Towards a standardized evaluation of imputation methodology

### 2022-08-01 Manuscript/Towards_a_standardized_evaluation_of_imputation_methodology.tex
Revisions (cont.)

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently separates the missing data problem from the analysis. When prediction is the goal, a single imputation may be sufficient \citet{sper20}. For inferential analyses, multiple imputation \citep{rubi76} has proven to be a valuable technique for obtaining valid inferences on incomplete datasets. In the case of multiple imputation each missing datum is imputed more than once, leading to multiple completed data sets. The resulting multiple inferences can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

@@

Standardizing the evaluation of imputation methodology requires simulators to consider common aspects in their simulation workflows. Table \ref{table:check} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed ADEMP structure (aims, data-generating mechanisms, estimands, methods, performance measures) for planning and reporting simulation studies, but highlight and add some specific recommendations for the evaluation of imputation methodology. [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

The ideal evaluation of imputation methodology depends on the aim of the specific study (i.e. how the evaluated methods are used in practice). A simulation study developed for the comparison of imputation methods will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. The suitability of the evaluations should be assessed with the imputation methods' purpose in mind, since the simulation aims and setup are naturally intertwined with the choice of imputation method(s) under evaluation.

We limit the scope of this paper to comparative simulation studies in the context of statistical inference. We thereby exclude the evaluation of imputation methodology for prediction and causal inference. Simulation study designs aimed at comparing different methods' predictive performance in incomplete data typically do not start out from a complete data set in which missingness is induced by the simulator \citep{liu21}. Rather, the methods are evaluated against the observed values of a certain outcome variable (or target) as the estimand. After imputing the missingness in the incomplete predictor (or feature) space, a prediction method is applied to estimate the outcome. Pairs of imputation and prediction methods with high predictive accuracy in one or more benchmark data sets are deemed as `good'. Note that such a simulation design does not have a `ground truth', so only the comparative performance of methods may be established. We do not recommend this approach if the inferential validity of the imputations is of interest. For an overview of missingness in the prediction context see e.g. \citet{sper20}. Missing data within a causal inference framework has been described by e.g. \cite{more18} and \cite{moha21}.

@@

% Before setting up a simulation study, the simulator should clearly define the scope of their evaluations. 
Estimand(s) or other simulation targets should be defined in the context of the study aim. [TODO: refer to \citet{pete14} about the need for a clear and unambiguous estimand.]

[TODO: describe problems with unclear simulation parameters, extrapolation beyond scope of simulations, QRPs \citep{pawe22} and misinterpretation \citep[see "investigator bias"][]{gree17}.]

[TODO: rewrite sampling variance paragraphs and combine with ``If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons."]

[TODO: add nuance because this doesn't always hold, see Tims review!] Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results. The comparative truth of the estimand(s) is determined by the choice of data-generating mechanism and whether sampling variance is included in the simulation setup, see Table \ref{table:dgm_truth}. If sampling variance cannot be omitted from the simulation scheme, multiple samples should be drawn from the data-generating mechanism (i.e., one sample per simulation repetition, a standard simulation setup). If sampling variance is not of interest, a single complete data set can be obtained from the data-generating mechanism (i.e., one sample for all simulation repetitions). This process is computationally convenient because only a single complete data set has to be considered during all of the simulations. We highly value the additional flexibility that such a simulation setup offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges.
When deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}.

\begin{table}[tb]
\begin{center}
\caption{Source of the comparative truth in simulation studies.}
\label{table:dgm_truth}
\begin{tabular}{lll}
\hline
               & With sampling variance      & Without sampling variance \\
               & (multiple samples drawn)    & (one sample drawn) \\
\hline  
Model-based simulation   & data-generating model         & single sample \\
Design-based simulation  & sufficiently large data set   & single sample \\
\hline
\end{tabular}
\end{center}
\end{table}

The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}.

@@

[TODO: add segway to reference method paragraph and add that performance on the complete data ofefrs an upper limit for performance.] In addition, the simulator should always perform complete case analysis. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating imputation performance. Complete case analysis may, therefore, serve as a benchmark method against which imputation performance should be evaluated. Moreover, pairing complete case analysis with MCAR and MAR mechanisms allow the simulator to evaluate the validity of the missing data generation process under the chosen analysis model.

>>>>

Imputation is a state-of-the-art technique for drawing valid conclusions from incomplete data. The technique has earned a permanent spot in research and policymaking, demonstrated e.g. by the detailed manual created by the National Research Council \citep*{little2012prevention}. Although top-down enforcement of valid ways to handle missing data is not yet very pronounced, an increasing amount of researchers and data scientists are embracing imputation techniques. After all, the principle of imputation is very intuitive. The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. The completed data that are thus obtained can be analyzed by standard techniques, which conveniently separates the missing data problem from the analysis.

For inferential analyses, multiple imputation \citep{rubi76} has proven to be a valuable technique for obtaining valid inferences on incomplete datasets. In the case of multiple imputation each missing datum is imputed more than once, leading to multiple completed data sets. The resulting multiple inferences can be combined into a single inference using Rubin's rules \citep[][p. 76]{rubi87}. When prediction is the goal of the analysis effort, a single imputation may be sufficient \citep{nijm20}. The quality of a solution obtained by imputation depends on the statistical properties of the incomplete data and the degree to which an imputation procedure is able to capture these properties when modeling missing values. In general, it holds that modeling missing data becomes more challenging when the amount of missingness increases. However, when (strong) relations in the data are present, the observed parts can hold great predictive power for the models that estimate the missingness. In that case, imputation would be substantially more efficient than the ubiquitous complete case analysis.

@@

Standardizing the evaluation of imputation methodology requires simulators to consider common aspects in their simulation workflows. Table \ref{table:check} outlines some suggested steps to adopt. Please note that there is--of course--no `silver bullet' and that we do not claim to present a universally applicable approach. Our suggestions are the accumulated result of scientific literature, research experience, and discussion. For an excellent overview of general best practices for method evaluation by means of simulation, see \citet{morr18}. We recommend adhering to their proposed structure for planning and reporting simulation studies and \citet{pawe22}'s advice to pre-register simulation protocols. We henceforth highlight and add some recommendations for the evaluation of imputation methodology. 
% [TODO: Add advice to pre-register simulation protocols, which allows for fair comparisons of per-protocol results \citep{pawe22}.]

@@

The ideal evaluation of imputation methodology is study specific. A simulation study developed for the comparison of several imputation methods will typically have a different design than one aimed at establishing the inferential validity of a single (novel) method. How the evaluated methods are used in practice thereby governs the simulation procedure. Since the simulation aims and setup are intertwined with the choice of imputation methods under evaluation, the suitability of the evaluations should be assessed with the imputation methods' purpose in mind.

We limit the scope of this paper to comparative simulation studies in the context of statistical inference. With that, we exclude the evaluation of imputation methodology for prediction or causal inference. The type of simulation study under consideration in this paper requires a form of `comparative truth' to assess the inferential validity of imputation methods. This is in contrast to simulation studies aimed at comparing the predictive performance of imputation and prediction method pairs. Such a simulation design typically does not start out from a `ground truth' (e.g., a complete data set in which missingness is subsequently induced by the simulator). Rather, the imputation and prediction method pairs are evaluated on their ability to yield high predictive accuracy in one or more incomplete benchmark data sets \citep{liu21}. In these designs, only the comparative performance of methods may be established. We do not recommend this approach if the inferential validity of the imputations is of interest. For an overview of missingness in the prediction context see e.g. \citet{wood15} and \citet{sper20}. Missing data within a causal inference framework has been described by e.g. \cite{more18} and \cite{moha21}.

@@

Before setting up a simulation study, the simulator should clearly define and describe the scope of their evaluations. If the simulation parameters are unclear, methodological problems with simulation study design may be obfuscated by the cognitive problem of generalization \citep{gree17}. For example, misinterpretation and (unintended) questionable research practices can prompt extrapolation beyond the scope of the simulations \citep{gree17, pawe22}. Imputation methods that seem to work well in simulations may then not be suitable in practice, when applied to incomplete empirical data.

One potential reason for discrepancies between imputation method performance in simulations versus applications is the use of sampling variance in the simulation design. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation \citep{vink14}. If sampling variance is not of interest, a single complete data set can serve as `comparative truth' in all simulation repetitions (as opposed to a `standard' simulation setup with a different complete data set per repetition). After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Removing sampling variance from performance evaluation is computationally convenient and may potentially sharpen comparisons between imputation methods, clarifying inconclusive results on their ability to correct for missingness. However, when deviating from the standard simulation workflow, conventional pooling rules for multiple imputation \citep[cf.][p. 76-77]{rubi87} do not apply and alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. This opens the door towards human error in the implementation and interpretation of such a simulation study. On the one hand, we highly value the additional flexibility that this design offers. Especially in the case of data transformations, it can be challenging to derive the true parametric references to evaluate against. Simply using a single generated complete set in which missingness is induced, and evaluating against that generated data set avoids a plethora of procedural problems and computational challenges. On the other hand, we want to warn simulators against potential mishaps, e.g. when calculating and comparing standard errors.

% [TODO: describe problems with unclear simulation parameters, extrapolation beyond scope of simulations, QRPs \citep{pawe22} and misinterpretation \citep[see "investigator bias"][]{gree17}.]

% Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

% The comparative truth of the estimand(s) is determined by the choice of data-generating mechanism and whether sampling variance is included in the simulation setup, see Table \ref{table:dgm_truth}.

% If sampling variance cannot be omitted from the simulation scheme, multiple samples should be drawn from the data-generating mechanism (i.e., one sample per simulation repetition, a standard simulation setup).

% If sampling variance is not of interest, a single complete data set can be obtained from the data-generating mechanism (i.e., one sample for all simulation repetitions).

% \begin{table}[tb]
% \begin{center}
% \caption{Source of the comparative truth in simulation studies.}
% \label{table:dgm_truth}
% \begin{tabular}{lll}
% \hline
%                & With sampling variance      & Without sampling variance \\
%                & (multiple samples drawn)    & (one sample drawn) \\
% \hline  
% Model-based simulation   & data-generating model         & single sample \\
% Design-based simulation  & sufficiently large data set   & single sample \\
% \hline
% \end{tabular}
% \end{center}
% \end{table}
%

[TODO: fix segway!] For fair evaluations, the following simulation design choices may be considered. Estimands and other simulation targets should be clearly and unambiguously defined in the context of the study aim \citet{pete14}. The simulation design should ideally be fully factorial, i.e. varying each simulation condition against all other conditions. There may, for example, be interactions between the source and amount of missingness in the incomplete data and the efficacy of imputation methods (i.e., the validity of the assumed missingness mechanism becomes increasingly important with higher missingness proportions). The number of simulation repetitions may be informed by the required level of precision in the simulation study \citep[e.g. as determined from a maximum tolerable level of uncertainty in terms of a performance measure's Monte Carlo error][]{morr18}. If there is more than one imputation method under evaluation, the simulation design should apply each method to every incomplete data set. Applying all methods to the same incomplete data set is computationally convenient, and minimizes unnecessary variation, which makes for fairer comparisons.

@@

[TODO: add segway to reference method paragraph and add that performance on the complete data offers an upper limit for performance.] In addition, the simulator should always perform complete case analysis. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating imputation performance. Complete case analysis may, therefore, serve as a benchmark method against which imputation performance should be evaluated. Moreover, pairing complete case analysis with MCAR and MAR mechanisms allow the simulator to evaluate the validity of the missing data generation process under the chosen analysis model.
